---
title: "1. Introduction to Rstudio and R"
author: "Dr. Uma Ravat"
date: "Feb 2024"
subtitle: '<br><br>Introduction to Social Network Analysis in R '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F}
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


# 1. Tour of four `Rstudio` panes

## -------~~ 1. Console/Terminal pane (bottom left) --------

The Console pane is at the bottom left. This is  where all calculations are performed

Try the following in the console

$$ 3 + 2 = ? \quad \quad log(10) = ? \quad \quad 32 =? $$
Were you able to do this in the console? 

`r bfcolor("Solutions:", "blue")`



## -------~~ 2. Scripting/Source code pane (top left)  --------

This is where writing of text and/ new functions/scripts/code should be done.


## -------~~ 3. Files/Plots/Packages/Help/Viewer pane (bottom right) --------

Can you find 'igraph' package? 

Can you install it? Note down below the command that installs the package. 

Can you load the package so that the data and functions are available for you to use? 

Verify in the "Packages" pane that `igraph` package has been loaded and then unload the `igraph` package

`r bfcolor("Solutions:", "blue")`

```{r}
# install command for installing igraph package


# command to load igraph package is library(igraph)


```


## -------~~ 4. Environment/History pane (top right) --------

- Can you see any commands in the History pane?
- What do you see in the Environment pane?

Environment pane shows what variables/Data/Functions are stored in your current Rsession.

`r bfcolor("Solutions:", "blue")`


That completes a Tour of `Rstudio` panes. If you have finished, help someone who is not yet done or ask any questions. 



# 2. Tour of  `R` programming language

## 2a. -------~~ Basics of navigating `R` programming language --------

## -------~~ Running code --------


- Run entire chunk using the "green triangle" 
- Run any part of your code by selecting it an pressing Ctrl+Enter / Cmd+Enter

```{r eval = FALSE}
3 + 2
log(10) 
log(10) =           
```

## -------~~ Comments in R--------

In a code chunk, any part of the code  starting with '#' is a comment 

## -------~~ Help in R--------

| Function | Action                                                                     |
|------------------------|----------------------------------------------------------------------------|
| ?foo     | Help on the function `foo`                                                 |
| ??foo    | Search the help system for instances of the function `foo`                 |
| data()   | List all available example datasets contained in currently loaded packages |
| help(nameOfDataset)   | provides you a detailed description of the dataset in question |
| getwd()  | List the current working directory                                         |
| setwd()  | change current working directory to new location specified in `setwd()`    |
| ls()     | List the objects in the current directory                                  |

## -------~~ Useful keyboard shortcuts --------

| Shortcut | Action                                                                     |
|----------------------------|----------------------------------------------------------------------------|
| [Tab]     | Autocomplete                                                 |
| [Cmd/Cntl + Enter]     | Run/Execute code                                                 |
| [Cmd/Cntl + Option + I]     | Add code chunk                                                |


## -------~~ John Chambers about R --------

> To understand computations in R, two slogans are helpful:

>   Everything that exists is an object.

>   Everything that happens is a function call.

> — John Chambers

Even a *function* is an *object*.


## 2b. -------~~ Basics of `R` language --------

## -------~~ Object Assignments and Evaluation --------


- You can assign a value to an object using `R`'s assignment operators : "<-", or "=".
- If you are just starting out, use "<-", there is less chance of confusion about what is assigned to what with the "<-" operator compared to others that `R` allows you to use.


```{r eval = FALSE}

x <- 2         # Assignment
x              # Evaluate the expression and print result

y <- 3         # Assignment
y + 5          # Evaluation, y remains 3

z <- x + 17*y  # Assignment
z              # Evaluation

rm(z)          # Remove z: deletes the object.
z              # Error!

# Notes on R
x <- 5         # use this
x = 5          # also works 
x -> 5         # though awkward it also works, but don't use. 

X <- 3        # R is case sensitive
```




## -------~~ Naming Objects --------


- You can give the objects you create more helpful names.
- They should start with a letter or a period, and contain letters, numbers, underscores, or periods.
- Use 'CamelCase' or '_' to improve readability

```{r}
num_friends <- 5
num_friends <- num_friends + 1
num_friends # Made a new friend!
```


##  -------~~ Value comparisons --------

- Comparisons return the logical constants TRUE or FALSE 

```{r}

x <- 2         # Assignment
x              # Evaluate the expression and print result

y <- 3         # Assignment
y + 5          # Evaluation, y remains 3


2==2  # Equality
2!=2  # Inequality; ! in R means "not"
x <= y # less than or equal: "<", ">", and ">=" also work


# You can force TRUE and FALSE to behave like numbers
# In that case, TRUE will be 1, and FALSE will be 0
TRUE + TRUE 
5*TRUE
5*FALSE 

# TRUE and FALSE can be abbreviated to T and F (capitalized!)
T==F
T>F
!T
# Try lowercase and see what happens
```

##  ------~~ Special constants -------- 

- NA, NULL, Inf, -Inf, NaN

```{r}
# NA - missing or undefined data
5 + NA      # When used in an expression, the result is generally NA
is.na(5+NA) # Check if missing?

# Note that is.na() above is an example of a function in R
# It uses the parameters listed in parentheses to compute and return a result


# NULL - an empty object, e.g. a null/empty list
10 + NULL     # use returns an empty object (length zero)
is.null(NULL) # check if NULL (is.null is another function)

# Inf and -Inf represent positive and negative infinity
# They can be returned by  mathematical operations like division of a number by zero:

5/0
is.finite(5/0) # Check if a number is finite

# NaN (Not a Number) - the result of an operation that cannot be reasonably defined 
0/0
is.nan(0/0)
```

##  ------~~ Vectors --------  

- Most basic and widely used data structure in R

```{r vectors}
# We can create a vector with c(), a function which combines values known as the combine function
v1 <- c(1, 5, 11, 33)       # Numeric vector, length 4
v2 <- c("hello","world")    # Character vector, length 2 (a vector of strings)
v3 <- c(TRUE, TRUE, FALSE)  # Logical vector, length 3, same as c(T, T, F)
```


```{r}
# Coercion in R 
# Combining different types of elements in one vector will coerce the elements 
# to the least restrictive type:

v4 <- c(v1,v2,v3,"boo") 	# All elements turn into strings
class(v4)
```

```{r}
# Other ways to create vectors:
v <- 1:7         # same as c(1,2,3,4,5,6,7) 
v
v <- rep(0, 77)  # repeat zero 77 times: v is a vector of 77 zeroes
v
v <- rep(1:3, times=2) # Repeat 1,2,3 twice  
v
v <- rep(1:10, each=2) # Repeat each element twice  
v
v <- seq(10,20,2) # sequence: numbers between 10 and 20, in jumps of 2 
v

length(v)        # check the length of the vector 
```

```{r}
# Vector operations: Vectorized operations in R make R efficient!
v1 <- 1:5         # 1,2,3,4,5
v2 <- rep(1,5)    # 1,1,1,1,1 

# Element-wise operations:
v1 + v2      # Element-wise addition
v1 + 1       # Add 1 to each element
v1 * 2       # Multiply each element by 2
v1 + c(1,2)  # Problem: c(1,2) is a vector of a different length; Warning and R still  complete your request by using "recycling"

# Mathematical operations on vectors:
sum(v1)      # The sum of all elements
mean(v1)     # The average of all elements
sd(v1)       # The standard deviation
min(v1)      # the minimum value
max(v1)      # the maximum value
cor(v1,v1*5) # Correlation between v1 and v1*5 

# Logical operations on vectors: These are also vectorized operations! Cool!

v1 > 2       # Each element is compared to 2, returns logical vector
v1==v2       # Are corresponding elements equivalent, returns logical vector.
v1!=v2       # Are corresponding elements *not* equivalent? Same as !(v1==v2)
(v1>2) | (v2>0)   # the | is the logical operator OR, returns a vector
(v1>2) & (v2>0)   # the & is the logical operator AND, returns a vector

all(v1>3) # Function all() checks if something is true for ALL elements
any(v1>3) # Function any() checks if something is true for ANY elements

# Extracting specific elements of a vector is called subsetting a vector
v1             # Let's recall what v1 is

# Vector elements: Index in R starts at 1
# NOTE: If you are used to languages indexing from 0, R will surprise you by indexing from 1.

v1[3]             # third element of v1
v1[2:4]           # elements 2, 3, 4 of v1
v1[c(1,3)]        # elements 1 and 3 - note that your indexes here are a vector
v1[c(T,T,F,F,F)]  # elements 1 and 2 - only returns the ones that are TRUE
v1[v1>3]          # v1>3 is a logical vector TRUE for elements >3; returns index and not value/element at that index
v1[-3]            # negative number: will return all elements *except* the third



# To replace elements in a vector, assign them a new value
v1[5] <- 10
v1

# To add more elements to a vector, simply assign them values.
v1[6:10] <- 6:10
v1

# We can even directly assign the vector a length:
length(v1) <- 15 # the last 5 elements are added as missing data: NA
v1
```


##  ------~~ Factors --------
```{r factors}
# Factors can be used to store categorical data.

home_state_v <- c("NJ", "CA", "NJ", "NY", "NY", "NY", "PA")         #vector
home_state_f <- factor(c("NJ", "CA", "NJ", "NY", "NY", "NY", "PA")) #factor
home_state_v
home_state_f

# R will identify the different levels of the factor - e.g. all distinct values. 
# The data is stored internally as integers - each number corresponding to a factor level.

levels(home_state_f)  # The levels (distinct values) of the factor (categorical variable)
levels(home_state_f) <- c("CA","NJ","NY","PA","TX","MA")
home_state_f

as.numeric(home_state_f)  # The factor as numeric values: 1 is CA, 2 is NJ, 3 is NY, 4 is PA
as.numeric(home_state_v)  # The character vector, however, can not be coerced to numeric values

as.character(home_state_f)  # convert factor to character vector
as.factor(home_state_v)     # convert vector to factor

# Sometimes, our categorical data can be ordered.
# For example, we may have a survey question asking students
# how they feel about the pace of material covered this week. 

weekpace_v <- c("very fast", "just right", "just right", "slow", "just right", "fast")
weekpace_levels <- c("very fast", "fast","just right", "slow", "very slow")


weekpace_f <- factor(weekpace_v, levels=weekpace_levels, ordered=TRUE)
weekpace_f

# In the code above, we did two things: mentioned a possible level
# that was not in our data (fast, very slow) and specified order
```


##  ------~~ Matrices & Arrays -------- 
```{r matricesArrays}
 # A matrix is a vector with dimensions:
m <- rep(1, 20)   # A vector of 20 elements, all 1
m

dim(m) <- c(5,4)  # Dimensions set to 5 & 4, so m is now a 5x4 matrix
m

# Create a matrix using matrix():
m <- matrix(data=1, nrow=5, ncol=4)  # same matrix as above, 5x4, full of 1s
m


m <- matrix(1,5,4) 			             # same matrix as above
m


dim(m)                               # What are the dimensions of m?


# Functions (such as matrix above) often require that we provide some parameters.
# By default, matrix is filled in by columns. 
# To check syntax run ?matrix in console
# Here, for example, we have 'data', 'nrow' (number of rows), and 'ncol' (number of columns).
# If we do not mention the parameter names, they are assumed based on the given order --
# the first thing in the parentheses is expected to be data, the second rows, the third columns.

# Create a matrix by combining vectors:
m <- cbind(1:5, 5:1, 5:9)  # Bind 3 vectors as columns, 5x3 matrix
m

m <- rbind(1:5, 5:1, 5:9)  # Bind 3 vectors as rows, 3x5 matrix
m

# If the data we provide has fewer than 10x10=100 elements, it gets repeated. In R this is called recycling
m <- matrix(1:10,10,10)  # data = 1:10, nrow = 10, ncol = 10  so 10x10 matrix with recycling
m

# Select matrix elements: Called subsetting in R:

m[2,3]  # Matrix m, row 2, column 3 - a single cell
m[2,]   # The whole second row of m as a vector
m[,2]   # The whole second column of m as a vector
m[1:2,4:6] # submatrix: rows 1 and 2, columns 4, 5 and 6
m[-1,]     # all rows *except* the first one

m[1,]==m[,1]  # Are elements in row 1 equivalent to corresponding elements from column 1? 
m>3           # A logical matrix: TRUE for m elements >3, FALSE otherwise
m[m>3]        # Selects only TRUE elements - that is ones greater than 3; searches by columns


t(m)          # Transpose m  (switch rows and columns)   
m <- t(m)     # Assign m the transposed m
m %*% t(m)    # %*% does matrix multiplication
m * m         # * does element-wise multiplication

# Arrays: more than 2 dimensions: indexed as row, column, layer
# Created with the array() function:

a <- array(data=1:18,dim=c(3,3,2)) # 3d with dimensions 3x3x2
a <- array(1:18,c(3,3,2))          # the same array
a
```


##  ------~~ Lists --------  
```{r lists}
# Lists are collections of objects (e.g. of strings, vectors, matrices, other lists, etc.)

l1 <- list(boo=v1,foo=v2,moo=v3,zoo="Animals!")  # A list with four components
l1

l2 <- list(v1,v2,v3,"Animals!")
l2


l3 <- list() # empty list
l3

l4 <- NULL   # null (empty) object
l4


l1["boo"]      # Access boo: this returns a list.
l1[["boo"]]    # Access boo: this returns the numeric vector
l1[[1]]        # Returns the first component of the list, equivalent to above.# Naming the components is helpful.
l1$boo         # Named elements can be accessed using the $ operator - equivalent to [[]]

# How can I tell if something is a list or not?
is.list(v1) # is this a list?
is.list(l1) # is this a list?
class(l1)   # what kind of object is this?

# Add more elements to a list:
l3
l3[[1]] <- 11 # add an element to the empty list l3, so l3 now has one component
l3

l4
l4[[3]] <- c(22, 23) # add a vector as component 3 in the empty list l4. 
l4

# Since we added element 3, elements 1 & 2 will be generated and empty (NULL)


l1[[5]] <- "More elements!" # The list l1 had 4 elements, we're adding a 5th here.
l1

l1[[8]] <- 1:11 # We added an 8th element, but not 6th or 7th. Those will be created empty (NULL)
l1

l1$Something <- "A thing"  # Adds a ninth element - "A thing", named "Something"
l1

```

##  ------~~ Data Frames --------  
```{r dataframes}

# The data frame is a special kind of list used for storing our dataset(from surveying students) as a table.
# Think of rows as cases(subjects/students), columns as variables(pace of material this week, how many hours they spent on homework etc). Each column is a vector or factor.

# Creating a dataframe:

dfr1 <- data.frame( ID=1:4,
                    FirstName=c("John","Jim","Jane","Jill"),
                    Female=c(F,F,T,T), 
                    Age=c(22,33,44,55) )
dfr1

dfr1$ID        # Access the ID column of dfr1. 
class(dfr1$ID) # What type is that column?

dfr1$FirstName        # Access the FirstName column of dfr1. 
class(dfr1$FirstName) # What type is that column?

# Access elements of the data frame
dfr1[1,]   # First row, all columns
dfr1[,1]   # First column, all rows
dfr1$Age   # Age column, all rows
dfr1[1:2,3:4] # Rows 1 and 2, columns 3 and 4 - the gender and age of John & Jim
dfr1[c(1,3),] # Rows 1 and 3, all columns

# Find the names of everyone over the age of 30 in the data
dfr1[dfr1$Age>30, 2]

# Find the average age of all females in the data:
mean ( dfr1[dfr1$Female==TRUE, 4] )

```

## 2c. -------~~ Essentials of `R` programming language --------

##  ------~~ Flow Control --------

```{r flowcontrol}

# if (condition) expr1 else expr2
x <- 5
y <- 10
if (x==0) y <- 0 else y <- y/x  
y

# To do several things if x equals 0, we can enclose the block of code
# that includes all of those things in {} like this:
if(x>0) {
  x <- 10
  y <- 42
}

# for (variable in sequence) expr
ASum <- 0
AProd <- 1
for (i in 1:x) {
  ASum <- ASum + i
  AProd <- AProd * i
}

ASum  # equivalent to sum(1:x)
AProd # equivalent to prod(1:x)

# while (condition) expr
x
while (x > 0) {
  print(x)
  x <- x-1 }

# repeat expr, use break to exit the loop
repeat { 
  print(x)
  x <- x+1
  if (x>10) break }
```

##  ------~~ R packages --------
```{r packages}
# Much of the functionality of R is contained in packages.
# Packages group together related code and data that we can use.
# Anyone can create a package.
# "Official" versions of packages are stored in the CRAN repository.
# You can install those directly as long as you know the package name.

# For example: 
#install.packages("igraph") 

# You can start using any package by loading it first:
library("igraph")          

# That allows you to access functions and data from that package.
# For example, graph.ring() is a function from the 'igraph' package.
# It creates a ring graph with a certain number of nodes (here 7)
g <- graph.ring(7)
plot(g)

# To get help on a package, try this:
package?igraph

# When you are done using a package, it is a good idea to let R know
# Detaching a package:
detach(package:igraph)    

```

##  ------~~ Read/write files --------
```{r rwfiles}
# Set the working folder where you want to read/write files:
# Note the forward slashes (as opposed to backslash in Windows paths)
#setwd("C:/Classes/SNA")
# Can do this in Rstudio using the Session menu

dfr1 <- data.frame( ID=1:4,
                    FirstName=c("John","Jim","Jane","Jill"),
                    Female=c(F,F,T,T), 
                    Age=c(22,33,44,55) )

# Save to a CSV file:
write.csv(dfr1, "my_data_frame.csv", row.names=F)

# Read the CSV file back:
dfr2 <- read.csv("my_data_frame.csv")
dfr2
```

#  ------~~ R troubleshooting --------

```{r troubleshoot}
# While coding, it is natural to  generate many (and often very creative) errors in R, there are three
# simple things that will most often go wrong. Those include: 

# 1) ~~Capitalization~~. R is case sensitive - a graph vertex named "Jack" is not the same
# as one named "jack". The function "rowSums" won't work as "rowsums" or "RowSums".
#
# 2) ~~Object class~~. While many functions are willing to take anything you throw
# at them, some will still surprisingly require character vector or a factor instead of
# a numeric vector, or a matrix instead of a data frame. Functions will also occasionally
# return results in an unexpected format.
#
# 3) ~~Package namespaces~~. Occasionally problems will arise when different packages
# contain functions with the same name. R may warn you about this by saying something
# like "The following object(s) are masked from 'package:igraph'" as you load a package.
# One way to deal with this is to call functions from a package explicitly using '::'.
# For instance, if function 'blah' is present in packages A and B, you can call
# A::blah and B::blah. In other cases the problem is more complicated, and you may
# have to load packages in certain order, or not use them together at all. 
 
# For more advanced troubleshooting, check out try(), tryCatch(), and debug().
?tryCatch

```





































