---
title: "3 WA. Descriptive Network Statistics (Metrics) "
author: "Dr. Uma Ravat"
date: "Feb 2024"
subtitle: '<br><br>Introduction to Social Network Analysis in R '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F}
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


# 1. Setting up the network 

Load the `UKFaculty` Network

```{r setup_network}
# (remove # below and run this if they are not already installed )
# once installed put # back to avoid warnings/errors
#install.packages("igraph")
#install.packages("igraphdata")

library(igraphdata)
library(igraph)

## Some setup so plots are bigger with no margins
#default margins
def_marg <- c(5.1, 4.1, 4.1, 2.1)

#no margins
no_marg <- c(0, 0, 0, 0)

## load the data
data(UKfaculty)
```

# 2. Descriptive Network Statistics (Metrics)

```{r}

# ================  ~~ Descriptive Network Statistics (Metrics) ================
# Density 
graph.density(UKfaculty)

# Reciprocity 
reciprocity(UKfaculty)

# Dyad Census 
dyad_census(UKfaculty)

# Transitivity 
transitivity(UKfaculty)

# Reachability 
average.path.length(UKfaculty)

diameter(UKfaculty, weights = NA)

#diameter(UKfaculty)

farthest_vertices(UKfaculty)

#farthest_vertices(UKfaculty, , weights = NA)

# Connected components
is.connected(UKfaculty, mode = "strong")

# Articulation points
articulation.points(UKfaculty)

```


# 3. Network visualization

### Basic plot

```{r}
#basic plot
plot(UKfaculty)
```



### Improving the plot

**Option 1:** Setting visual properties outside the plot function

We add attributes to the network

**Step 1** Set vertex and/or edge attributes

```{r}

# vertex properties 
V(UKfaculty)$color <- "#88398A"

# vertex label properties
V(UKfaculty)$label.color <- "white"
V(UKfaculty)$label.cex <- 0.5

# edge properties
E(UKfaculty)$color <- "gray"

```

**Step 2** Plot network

```{r}

plot(UKfaculty)

```

### Option 2 Setting visual properties inside the plot function

We use the function arguments

```{r}

plot(UKfaculty, 
     # === vertex properties 
     vertex.color = "#88398A",
     vertex.frame.color = "#88398A",
     vertex.size = 9,
     # === vertex label properties
     vertex.label.cex = 0.5,
     vertex.label.color = "white",
     vertex.label.family = "Helvetica",
     # === edge properties
     edge.color = "gray",
     edge.width = 0.5,
     edge.arrow.size = 0.2)
```


This helps us add different visual properties to subsets of edges or nodes. 

### Coloring nodes by group to which they belong

```{r}
table(V(UKfaculty)$Group)

mycolors <- c("black", 
              "#D3D3D3", 
              "#88398A", 
              "white", 
              "#562457")
mylabcols <- c("white", "black", "white", "black", "white")

V(UKfaculty)$color <- mycolors[V(UKfaculty)$Group]
V(UKfaculty)$frame.color <- mycolors[V(UKfaculty)$Group]
V(UKfaculty)$label.color <- mylabcols[V(UKfaculty)$Group]

plot(UKfaculty)
```

We can also combine both options. This is usually what we do. 

```{r}
plot(UKfaculty,
     vertex.size = 9,
     # === vertex label properties
     vertex.label.family = "Helvetica",
     # === edge properties
     edge.width = 0.5,
     edge.arrow.size = 0.1)
```


### Coloring specifc edges

```{r}
E(UKfaculty)$color <- "gray"
E(UKfaculty)$color[E(UKfaculty)[47%--%28]] <- "red"

plot(UKfaculty, 
     # === vertex properties 
     vertex.size = 9,
     # === vertex label properties
     vertex.label.cex = 0.5,
     vertex.label.family = "Helvetica",
     # === edge properties
     edge.width = 0.5,
     edge.arrow.size = 0.1)

```




# 4. Centrality

## Degree Centrality 

```{r}

# Indegree 
degree(UKfaculty, mode="in")
par(mar= def_marg)
hist(degree(UKfaculty, mode = "in"),
     breaks=1:vcount(UKfaculty)-1, 
     main="Indegree histogram")

# Outdegree
degree(UKfaculty, mode="out")
hist(degree(UKfaculty, mode = "out"),
     breaks=1:vcount(UKfaculty)-1, 
     main="Outdegree histogram")

# Combine indegree and outdegree 

deg <- degree(UKfaculty, mode="all")

V(UKfaculty)$deg <- deg

par(mar= no_marg)

plot(UKfaculty, 
     vertex.size=deg)

plot(UKfaculty, 
     vertex.size=sqrt(deg)+1,
     vertex.label.cex = 0.4,
     edge.arrow.size = 0.1)

# arrange by degree
sort(deg, decreasing = T, index.return = T)

# uniform coloring
V(UKfaculty)$color <- "white"
V(UKfaculty)$label.color <- "black"
V(UKfaculty)$frame.color <- "black"

#coloring 5 most central nodes
V(UKfaculty)$color[V(UKfaculty)[c(29,37,77,62,52)]] <- "#88398A"

plot(UKfaculty, 
     vertex.size=10,
     vertex.label.cex = 0.4,
     edge.arrow.size = 0.1)
```

## Closeness Centrality 

```{r}
cl <- closeness(UKfaculty, mode="all", weights=NA) 

centr_clo(UKfaculty, mode="all", normalized=T)

# arrange by closeness
sort(cl, decreasing = T, index.return = T)

# uniform coloring
V(UKfaculty)$color <- "white"

#coloring 5 most central nodes
V(UKfaculty)$color[V(UKfaculty)[c(29, 37, 62, 52, 69)]] <- "#88398A"

plot(UKfaculty, 
     vertex.size=10,
     vertex.label.cex = 0.4,
     edge.arrow.size = 0.1)
```

## Betweenness Centrality 

```{r}
btw <- betweenness(UKfaculty, directed=T, weights=NA)

edge_betweenness(UKfaculty, directed=T, weights=NA)

centr_betw(UKfaculty, directed=T, normalized=T)

# sorting by betweenness 
sort(btw, decreasing = T, index.return = T)

V(UKfaculty)$color[V(UKfaculty)[c(62, 29, 37, 38, 52)]] <- "#88398A"

plot(UKfaculty, 
     vertex.size=10,
     vertex.label.cex = 0.4,
     edge.arrow.size = 0.1)
```

# 5. Your Turn: Descriptive statistics for the karate network

Obtain measures of density, reciprocity, dyad census, transitivity, average path length, diameter, and articulation points for the karate network. 

_Be sure to load the network first_

_Add code chunks as necessary to run your code and answer the following_

```{r}

```


`r bfcolor("Question:", "blue")` Based on these metrics, what can you say about this network?

