---
title: "2 WA. Introduction to Networks"
author: "Dr. Uma Ravat"
date: "Feb 2024"
subtitle: '<br><br>Introduction to Social Network Analysis in R '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F}
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


# 1. Introduction to Networks

## -------~~ 1.Set up --------

```{r load_packages}
# (remove # below and run this if they are not already installed )
# once installed put # back to avoid warnings/errors
#install.packages("igraph")
#install.packages("igraphdata")

library(igraphdata)
library(igraph)

```

`r bfcolor("Question:", "blue")` What is the command to load a dataset? Load a library

`r bfcolor("Solution:", "blue")` 


Some setup so plots are bigger with no margins

``` {r}

#default margins
def_marg <- c(5.1, 4.1, 4.1, 2.1)

#no margins
no_marg <- c(0, 0, 0, 0)

```

## -------~~ 2.Load the data  --------

```{r}
data(UKfaculty)
```

`r bfcolor("Question:", "blue")` What is the command to load a dataset? Load a package?

`r bfcolor("Solution:", "blue")` 



## -------~~ 3.Plot of our network  --------

```{r}
par(mar=no_marg)
plot(UKfaculty, 
     vertex.size = 9,
     vertex.color = "#88398A",
     vertex.frame.color= "#88398A",
     vertex.label.family = "Helvetica", 
     vertex.label.cex = 0.5,
     vertex.label.color = "white",
     edge.arrow.size = 0.2)
```

# 2. Exploring our Network

##  Take a look first!

`r bfcolor("Question:", "blue")` What are the most general characteristics of this network?

```{r}
# ================  ~~ Exploring the  network ================
UKfaculty

# The description of the igraph object starts with up to four letters:
# 1. D or U, for a directed or undirected graph
# 2. N for a named graph (where nodes have a name attribute)
# 3. W for a weighted graph (where edges have a weight attribute)
# 4. B for a bipartite (two-mode) graph (where nodes have a type attribute)
#
# The two numbers that follow refer to the number of nodes and edges in the graph. 
# The description also lists graph, node & edge attributes, for example:
# (g/c) - graph-level character attribute
# (v/c) - vertex-level character attribute
# (e/n) - edge-level numeric attribute
# .... etc.
```

`r bfcolor("Question:", "blue")` Does this graph have multiple edges per pair of nodes? Does the graph have any loops?

_Recall: A graph is a simple graph if it does not contain loop edges and multiple edges._

```{r}
is.simple(UKfaculty)
```


# 3. Your Turn: Explore the basic characteristics of the karate network

Add code chunks as necessary to run your code and answer the questions below

- `r bfcolor("Question:", "blue")` How many nodes does the network have?
- `r bfcolor("Question:", "blue")` Is the network directed or undirected?
- `r bfcolor("Question:", "blue")` What vertex and edge attributes does the network have?
+ `r bfcolor("Question:", "blue")` Does the graph have multiple edges per pair of nodes or any loops?


# 4. Taking a closer look: Network Attributes 

## Graph Attributes

```{r}
graph.attributes(UKfaculty)
```

## Vertex Attributes

```{r vatrributes}
# ================  ~~  Vertex attributes  ================
 
vertex.attributes(UKfaculty)

V(UKfaculty)[[1:5]]

V(UKfaculty)$Group

table(V(UKfaculty)$Group)
```

## Edge Attributes

```{r eatrributes}
# ================  ~~  Edge attributes  ================
 
edge.attributes(UKfaculty)

E(UKfaculty)[[1:5]]

E(UKfaculty)$weight

summary(E(UKfaculty)$weight)

par(mar=def_marg)
hist(E(UKfaculty)$weight)
```

# 5. Accessing and Exploring Specific Elements 

## Exploring vertices 

### All vertices

```{r}
V(UKfaculty)
```

### Specific node connections 

`r bfcolor("Question:", "blue")` Which nodes are connected to node 1?

```{r}

# outgoing ties
neighbors(UKfaculty, 1, mode = "out")

#incoming ties
neighbors(UKfaculty, 1, mode = "in")

# all ties
neighbors(UKfaculty, 1, mode = "all")

```

`r bfcolor("Question:", "blue")` Which nodes have less than 10 connections?

```{r}
V(UKfaculty)[degree(UKfaculty) < 10]
```

## Exploring edges

`r bfcolor("Question:", "blue")` Does a given edge exist?

```{r}

UKfaculty[1,2]
UKfaculty[1,44]
UKfaculty[44,1]

```

`r bfcolor("Question:", "blue")` Edges between specific nodes

```{r}
E(UKfaculty)[1 %--% 2]

# regardless of the direction
E(UKfaculty)[1 %--% 44]

# from 44 to 1
E(UKfaculty)[1 %<-% 44]

# from 1 to 44
E(UKfaculty)[1 %->% 44]
```

`r bfcolor("Question:", "blue")` Edges containing specific vertices

```{r}

# All incoming edges of a vertex
E(UKfaculty)[to(1)]

# All outgoing edges of a vertex
E(UKfaculty)[.from(1)]

```
# 6. Your turn! Take a closer look at Karate network!

```{r}
data(karate)
```
Running this code will tell us the general characteristics of the network and the name of its attributes. 

```{r}
karate
```

**To answer each question in code chunks below**,
1. Replace every [xxx] with your code
2. Run the code chunk individually by clicking the "green triangle"
3. To see your solution in the knitted document, make sure to replace `eval=FALSE` with `eval=TRUE`

`r bfcolor("Question:", "blue")` How many factions does the karate network have? 
`r bfcolor("Question:", "blue")` How many nodes belong to each faction?

```{r, eval=FALSE}
table(V(karate)$[xxx])
```

`r bfcolor("Question:", "blue")` Which nodes nominated Mr. Hi?

```{r, eval=FALSE}

neighbors(karate, "Mr Hi", mode = [xxx])

```

`r bfcolor("Question:", "blue")` Which nodes nominated John A?

```{r, , eval=FALSE}
[xxx](karate, "John A", mode = "in")
```

`r bfcolor("Question:", "blue")` How many nodes nominated both Mr Hi and John A?

```{r, , eval=FALSE}

mh <- neighbors([xxx], "Mr Hi", mode = [xxx])
ja <- neighbors([xxx], "John A", mode = [xxx])

sum(mh %in% ja)

```

`r bfcolor("Question:", "blue")` Is there an edge between Mr Hi and John A?

```{r, eval=FALSE}
E(karate)["Mr Hi" [xxx] "John A"]
```

